---
title: "Class 5 Data Visualization with ggplot2"
author: "Barry (PID: 911)"
format: pdf
---

## Using GGPLOT

The ggplot2 package needs to be installed as it does not come with R "out of the box".

We use the `install.packages()` function to do this.

```{r}
head(cars)
```

To use ggplot I need to load it up before I can call any of the functions in the package. I do this with the `library()` function.

```{r}
library(ggplot2)
ggplot()
```

All ggplot figures have at least 3 things:
- data (the stuff we want to plot)
- aesthetic mapping (aes vales)
- geoms

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm")
```

ggplot is not the only graphing system in R there are lots of others. There is even "base R" graphics.


```{r}
plot(cars)
```


# Expression dataset


```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, color=State) +
  geom_point()
```


```{r}
sum(genes$State == "up")
```

```{r}
head(genes$State)
head(genes$State == "up")
```

```{r}
#install.packages("dplyr")

library(dplyr)
```



```{r}
head(genes)
```

```{r}
sum(genes$State == "up")
```


