---
title: "Class 07: Machine Learning 1"
author: "Barry (PID: 911)"
format: pdf
---

# Clustering

We will start today's lab with clustering methods, in particular so-called K-means. The main function for this in R is `kmeans()`.

Let's try it on some made up data where we know what thwe answer should be.

```{r}
x <- rnorm(10000, mean=3)
hist(x)
```

60 points
```{r}
tmp <- c(rnorm(30, mean=3), rnorm(30, -3) )
x <- cbind(x=tmp, y=rev(tmp))
head(x)
```

We can pass this to the base R `plot()` function for a quick.

```{r}
plot(x)
```

```{r}
k <- kmeans(x, centers = 2, nstart=20)
k
```

> Q1. How many points are in each cluster?

```{r}
k$size
```

> Q2. Cluster membership?

```{r}
k$cluster
```


> Q3. Cluster centers?

```{r}
k$centers
```

> Q4. Plot my clustering results

```{r}
plot(x, col=k$cluster, pch=16 )
```


> Q5. Cluster the data again with kmeans() into 4 groups and plot the results.

```{r}
k4 <- kmeans(x, centers = 4, nstart=20)
plot(x, col=k4$cluster, pch=16)
```

K-means is very popular mostly because it is fast and relatively straightforward to run and understand. It has a big limitation in that you need to tell it how many groups (k, or centers) you want.



# Hierarchical clustering

The main function in base R is called `hclust()`. You have to pass it in a "distance matrix" not just your input data.

You can generate a distance matrix with the `dist()` function.

```{r}
hc <- hclust( dist(x) )
hc
```

```{r}
plot(hc)
```

To find the clusters (cluster membership vector) from a `hclust()` result we can "cut" the tree at a certain height that we like. For this we use the `cutree()` function.

```{r}
plot(hc)
abline(h=8, col="red")
grps <- cutree(hc, h=8)
```


```{r}
table(grps)
```

> Q6. Plot our hclust results.

```{r}
plot(x, col=grps, pch=16)
```

# Principal Component Analysis


## PCA of UK food data

Read data showing the consumption in grams (per person, per week) of 17 different types of food-stuff measured and averaged in the four countries of the United Kingdom.

Let's see how PCA can help us but first we can try conventional analysis.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

I need to fix that first column...

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
```


A pairs plot can be useful if we don't have too many dimensions...

```{r}
pairs(x, col=rainbow(17), pch=16, cex=2)

```

## Principal Component Analysis (PCA)

PCA can help us make sense of these types of datasets. Let's see how it works.

The main function in "base" R is called `prcomp()`. In this case we want to first take the transpose `t()` of our input `x` so the columns are the food types and the countries are the rows.

```{r}
head( t(x) )
```

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

```{r}
pca$x
```

```{r}
plot( pca$x[,1], pca$x[,2], 
      col=c("orange", "red", "blue", "darkgreen"), 
      pch=16 )
```
The "loadings" tell us how much the origional variables (in our case the foods) contribute to the new variables i.e. the PCs

```{r}
head(pca$rotation)

## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )

```

Stop at Q9 


## ggplot

```{r}
library(tidyr)

# Add the rownames back as a column
y <- cbind(food=rownames(x),x)
rownames(y) <- NULL
head(y)
```

```{r}
z <- pivot_longer(y, 
                  cols = !food, 
                  names_to = "country", 
                  values_to = "count")
```

The first argument is the dataset to reshape, `y`

`cols` describes which columns need to be reshaped. In this case, it’s every column apart from `food`

`names_to` gives the name of the variable that will be created from the data stored in the column names, i.e. `country`.

values_to gives the name of the variable that will be created from the data stored in the cell value, i.e. count.




```{r}
library(ggplot2)

ggplot(z) +
  aes(food, count, col=country) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))
```

```{r}
ggplot(z) +
  aes(food, count) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))

```

```{r}
ggplot(z) +
  aes(food, count, fill=country) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))

```



```{r}
ggplot(z) +
  aes(food, count, fill=country) +
  geom_col(position='dodge') +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))

```

```{r}
ggplot(z) +
  aes(country, count, fill=food) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))

```

```{r}
ggplot(z) +
  aes(food, count, fill=country) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1)) +
  facet_wrap(vars(country), ncol=1)


```

## Using ggplot for these figures

We could use **ggplot** to make somewhat better figures than all of the above "base" R `plots()` and `barplots()`. Even though they may look better they won't really help us understand major trends to any better degree (we will do that with PCA in the next section).


```{r}
library(tidyr)
library(ggplot2)


# try with row.names
#m <- read.csv(url, row.names = 1)

q <- tibble::rownames_to_column(x, "Food")

# Put in longer format
foods <- pivot_longer(q, 
                  cols = !Food, 
                  names_to = "Country", 
                  values_to = "Count")

head(foods)

```


Now we can do some plotting

```{r}
ggplot(foods) +
  aes(Count, Food, col=Country) +
  geom_point()
```
```{r}
ggplot(foods) +
  aes(Count, Food, fill=Country) +
  geom_col()

```
```{r}
ggplot(foods) +
  aes(Count, Food, fill=Country) +
  geom_col(position='dodge') +
  theme_bw() +
  scale_x_continuous(expand=c(0,0))

```


```{r}
ggplot(foods) +
  aes(Count, Food, col=Country) +
  geom_point() +
  geom_segment(aes(x=0,
                   xend=Count,
                   y=Food, 
                   yend=Food), color="gray") +
  geom_point()
```

```{r}
ggplot(foods) +
  aes(Count, Food) +
  geom_boxplot() +
  geom_point(aes(col=Country))
```

```{r}
pairs(x, col=rainbow(17), pch=16)
```

```{r}
library(GGally)

ggpairs(x) 
```




## Optional: Using ggplot for these figures

We could use the **ggplot2** package to make somewhat better figures than all of the above "base" R `plots()` and `barplots()`. Recall that ggplot works with data.frames and unfortunately most of the output of these older base R functions like `prcomp()` are lists of vectors and matrices. 

So first we will need to take whatever it is we want to plot and convert it to a data.frame with the `as.data.frame()` function. Then to make our plotting life easier we will also add the food labels as a column (called "Food") to this data frame with the `rownames_to_column()` function from the **tibble** package (you might need to install this):




```{r}
df <- as.data.frame(pca$x)
df_lab <- tibble::rownames_to_column(df, "Country")

# Our first basic plot
ggplot(df_lab) + 
  aes(PC1, PC2, col=Country) + 
  geom_point()
```

And then we can get carried away and make this look much nicer:

```{r}
ggplot(df_lab) + 
  aes(PC1, PC2, col=Country, label=Country) + 
  geom_hline(yintercept = 0, col="gray") +
  geom_vline(xintercept = 0, col="gray") +
  geom_point(show.legend = FALSE) +
  geom_label(hjust=1, nudge_x = -10, show.legend = FALSE) +
  expand_limits(x = c(-300,500)) +
  xlab("PC1 (67.4%)") +
  ylab("PC2 (28%)") +
  theme_bw()
```


Let's do the same for our loadings/PC contributions figures. This data is stored in the `pca$rotation` object that we convert to a data frame, add the useful row names as a new column and then plot and customize with additional ggplot layers. Which do you prefer, base graphics or ggplot?


```{r}
ld <- as.data.frame(pca$rotation)
ld_lab <- tibble::rownames_to_column(ld, "Food")

ggplot(ld_lab) +
  aes(PC1, Food) +
  geom_col() 
```
We can now add some additional features to the plot, such as reordering the y axis by the PC1 loadings and selecting a rather ugly color scale (to match our country colors) and our prefered theme layer. 

```{r}
ggplot(ld_lab) +
  aes(PC1, reorder(Food, PC1), bg=PC1) +
  geom_col() + 
  xlab("PC1 Loadings/Contributions") +
  ylab("Food Group") +
  scale_fill_gradient2(low="purple", mid="gray", high="darkgreen", guide=NULL) +
  theme_bw()
```









